DisReserv (v1.0): Analytical displacement solution due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes

### 2020-09-11 manuscript/manuscript-reservoir.tex
reorganize files in repository
alterations codes and  manuscript

\title{DisReserv v1.0: Analytical displacement solution due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes}

@@

\Author[1]{Vanderlei}{C. Oliveira Jr.}
\Author[1]{Valeria}{C. F. Barbosa}
\Author[1,2]{Andre}{D. Arelaro}

@@

\correspondence{Vanderlei C. Oliveira Jr. (vandscoelho@gmail.com)}

@@

\runningauthor{Oliveira Jr., V. C., Barbosa, V. C. F. and Arelaro, A. D.}

@@

%%\twocolumn
\begin{abstract}
We have presented analytical solutions for the displacement due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes.
These solutions are based on the similarity between the gravitational potential yielded by  a volume source under a density variation and the displacement field yielded by a volume source in a half-space under a pressure variation.
This similarity enables the use of closed expressions of the gravitational potential and its derivatives for calculating the displacement field due to a volume source under a pressure variation.
We discretized the reservoir as a grid of 3D right rectangular prisms  juxtaposed in the horizontal and vertical directions.
Each grid prism has homogeneous pressure; however, pressure variations among different prisms are allowed. 
This parametrization of the reservoir yields a piecewise-constant distribution of pressure  in the subsurface. 
The discrete reservoir modeling to calculate the displacement field due to this pressure variation  follows the nucleus of strain concept in which the center of each prism represents the coordinate of a nucleus of strain.
The displacement due to a nucleus of strain is considered the infinitesimal element of the 
displacement due to an infinitesimal reservoir.
The displacement field due to the pressure of each prism is calculated by integrating the infinitesimal element of the displacement over the volume of the prism.
Finally, the displacement due to the reservoir can be approximated by the sum of the contributions of each prism of the discretized model.
We provide python codes (DisReserv) to calculate the displacement fields due to a reservoir with arbitrary shape and distribution of pressure changes. 
The displacement field is calculated inside and outside the reservoir. 
We calculate the displacement fields in three scenarios: i) cylindrical reservoir under uniform depletion; ii) cylindrical reservoir under non uniform depletion and iii) reservoir with arbitrary geometry and under arbitrary pressure distribution.
By calculating the  stress field at the free surface, we verify that the zero stress condition is satisfied.
\end{abstract}

\copyrightstatement{TEXT}

@@

The possibility of occurrence of catastrophic events, related to subsidence due to extraction or injection of fluids in reservoir, has stimulated efforts to  develop analytical methods for modeling the displacement and stress fields due to reservoir compactation.
The physical foundation of the displacement, stress and strain fields in the subsurface due to a reduction of pressure in the reservoir comes from the theory of thermoelasticity.
Theory of thermoelasticity has been laid in the first half of the nineteenth century to describe the interaction between the thermal field and elastic bodies.
In the uncoupled thermoelasticity theory for quasi-static problems (i.e., problems with negligible inertia effects), \cite{Goodier37} employed the method of superposition using
displacement potential functions and introduced the  concept of nucleus of thermoelastic strain in an infinite space.
Specifically, Goodier’s \citeyear{Goodier37} method simplified the thermoelastic problem by replacing it by an isothermal elastic problem with different boundary conditions together with the solution of a Poisson’s equation \citep{Tao71}. 
\cite{Mindlin&Cheng50} and \cite{Sen51} extended the Goodier's method to a homogenous half-space.
\cite{Sharma56} deduced the displacement and stress fields in an infinite elastic plate due to a nucleus of thermoelastic strain located at a point inside it by using infinite integrals involving Bessel functions.

@@

\cite{Geertsma&Opstal73} applied the nucleus of strain concept in the half-space 
to calculate the spatial subsidence distribution due to the production of reservoir with an arbitrary 3D shape.
By assuming a producing reservoir embedded in an a homogeneous, isotropic, and elastic medium, and a reservoir model in which the pressure perturbations are related to the displacement field by a linear relationship, \cite{Geertsma&Opstal73} discretized the reservoir into a grid of  prisms and calculated the displacement due to the pressure change in the whole reservoir  by the superposition of the displacement due to the constant pressure change in each prism.
\cite{Tempone10} adopted the same reservoir model used in \cite{Geertsma&Opstal73} and 
extended the nucleus of strain concept in the half-space to consider the effects of a rigid basement. 
Similarly, \cite{Tempone10} assumed a reservoir embedded in an a homogeneous, isotropic, and elastic medium and calculated the displacement, stress and strain fields subject to uniform depletion.
The main drawbacks in \cite{Geertsma&Opstal73} and \cite{Tempone10} are the assumption of homogeneous reservoir and the solution is only valid outside the reservoir.
In these case, the displacements within the reservoir are calculated by a linear interpolation of the displacements at the upper and lower edges of the reservoir \citep{Tempone12}.

Considering an inhomogeneous poroelastic model consists of layered stratigraphy, \cite{Mehrabian&Abousleiman15} developed closed-form formulae for the  displacement and stress  fields outside and inside of the reservoir embedded within elastic strata with different mechanical properties and subjected to pore pressure disturbances due to fluid extraction or injection. 
By assuming a linear elastic semi-infinite medium, \cite{Munoz&Roehl17} developed analytical solution for the  displacement field  outside and inside of an arbitrarily-shaped reservoir under arbitrary distribution of pressure changes. 
\cite{Munoz&Roehl17} parametrized the reservoir into a grid of 3D prisms and used the nucleus of strain concept. 
The nucleus of strain is taking as an infinitesimal volume element  for each prism and the displacement solution due to each prism is obtained by a three-dimensional integration over the prism volume. 
The displacement field outside and inside of the reservoir is given by the summation of the displacement fields of the displacements produced by all prisms setting up the reservoir model.

The present work assumes a linear elastic semi-infinite medium and provides an analytical solution for displacement field due to an arbitrarily-shaped reservoir under arbitrary distribution of pressure changes. 
Like \cite{Munoz&Roehl17} we used the nucleus of strain concept and discretized the
reservoir into a grid of 3D prisms along the $x-$, $y-$ and $-z$directions.
We also consider the nucleus of strain as an infinitesimal volume element  for each prism and the displacement solution due to each prism is obtained by a three-dimensional integration over the prism volume. 
The final displacement field due to the whole reservoir is the sum of the displacements produced by the prisms. 
In contrast with \cite{Munoz&Roehl17}'s method we take advantage the similarity between 
the  equations for calculating the displacement field due to a volume source in a half-space under a pressure variation and the gravitational potential due to a volume source under a density variation.
This similarity makes possible the use of closed expressions of the gravitational potential and its derivatives produced by the 3D right rectangular prism derived by \cite{Nagyetal2000} and (\citeyear{Nagyetal2002}) and \cite{Fukushima2020}  for calculating the displacement field due to a volume source under a pressure variation.
The adopted exact analytical formulae of the gravitational field  are valid expressions either outside or inside the prisms because the implemented expressions make use of  modified arctangent function proposed by \cite{Fukushima2020}. 
We present routines written in Python language (Python 3.7.6)  to calculate the displacement fields due to a reservoir with arbitrary shape and distribution of pressure changes. 
We validate our equations by verifying that at the free surface the stress fields are null.
Tests with synthetic data validate our approach.

@@

Here, we use a Cartesian coordinate system with the $x-$axis pointing to north, the $y-$axis pointing to east and the $z-$axis pointing downward.
By considering the discrete form of equation \ref{eq:displacement-solution}, the displacement field  $ {\bf{u}}_{i} \equiv  {\bf{u}}(x_{i}, y_{i}, z_{i})$ at an arbitrary point $(x_{i}, y_{i}, z_{i})$ due to the $j$th nucleus of strain at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ will be calculated by
\begin{equation}
{\bf{u}}_{i}  = {\bf{\nabla}} \: 
{\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \: + \: {\bf{{\nabla}_{2}}} {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}, 
\label{eq:displacement_ui}
\end{equation}
In equation \ref{eq:displacement_ui}, the functions 
$\phi_{1} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}$ and  
$\phi_{2} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}$
are, respectively, given by 
\begin{equation}
\phi_{1} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_1}_{ij} }
\label{eq:nucleus-phi1}
\end{equation}
and
\begin{equation}
\phi_{2} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_2}_{ij} }.
\label{eq:nucleus-phi2}
\end{equation}

@@

In equation \ref{eq:nucleus-phi1}, $ {R_1}_{ij}$ is the distance from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th coordinate of the nucleus of strain $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$, i.e.:
\begin{equation}
{R_1}_{ij} = {\sqrt{(x_{i}- x^{\prime}_{j})^{2} + (y_{i} - y^{\prime}_{j})^{2} + 
(z_{i} - z^{\prime}_{j})^{2}}}.
\label{eq:R1}
\end{equation}
In equation \ref{eq:nucleus-phi2}, $ {R_2}_{ij}$ is the distance from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th coordinate of the image nucleus $(x^{\prime}_{j}, y^{\prime}_{j}, - z^{\prime}_{j})$, i.e.:
\begin{equation}
{R_2}_{ij} = {\sqrt{(x_{i}- x^{\prime}_{j})^{2} + (y_{i} - y^{\prime}_{j})^{2} + 
(z_{i} + z^{\prime}_{j})^{2}}}.
\label{eq:R2}
\end{equation}
Figure \ref{fig:nucleus_strain} shows a schematic representation of the geometry
of the nucleus of strain problem in a semi-infinite medium. 
The $j$th nucleus of strain is located at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$. 
The $j$th image nucleus is located at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, - z^{\prime}_{j})$.
The distances from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th nucleus of strain and to the $j$th image nucleus are, respectively,  
$ {R_1}_{ij}$ (equation \ref{eq:R1})  and $ {R_2}_{ij}$ (equation \ref{eq:R2}) 
The free surface is a horizontal plane where the components of the stress are null.

Following the discrete form of the displacement solution (equation \ref{eq:displacement_ui}), 
the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus at the coordinates $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ can be written as:
\begin{equation}
{\bf{u}}_{i} (x_i,y_i,z_i) = {\bf{u_{1}}}_{i} (x_i,y_i,z_i) + {\bf{u_{2}}}_{i} (x_i,y_i,z_i), 
\label{eq:nucleus-ui_total}
\end{equation}
where ${\bf{u_{1}}}_{i} (x_i,y_i,z_i) \equiv {\bf{u_{1}}}_{i} $ is the gradient of the function $\phi_1$ (equation \ref{eq:nucleus-phi1}) 
\begin{equation}
{\bf{u_{1}}}_{i}  = {\bf{\nabla}} \: {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \:
= \frac{A  \: (1 + \nu)}{E} \: {\bf{\nabla}} \bigg( {\frac{1}{{R_1}_{ij}}} \bigg) 
\: \Delta p_{j} \: \:dv_j
\label{eq:nucleus-u1}
\end{equation}
and ${\bf{u_{2}}}_{i} (x_i,y_i,z_i) \equiv {\bf{u_{2}}}_{i} $ is obtained by  applying the operator ${\bf{{\nabla}_{2}}}$ (equation \ref{eq:nabla2}) to the imagem potential $\phi_2$ (equation \ref{eq:nucleus-phi2})
\begin{equation}
{\bf{u_{2}}}_{i} = {\bf{{\nabla}_{2}}} \: {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \:  
= \frac{A  \: (1 + \nu)}{E} \: 
\bigg[ 
(3 - 4\nu){\bf{\nabla}} \bigg(  {\frac{1}{{R_2}_{ij}}} \bigg)   \:  
+ 2 {\bf{\nabla}} \bigg( z \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}} \bigg) \: 
- 4(1- \nu){\bf{\hat{z}}} {\bf{{\nabla}^{2}}}  \bigg( {\frac{z}{{R_2}_{ij}}} \bigg)
\bigg]   \: \Delta p_{j} \: \:dv_j , 
\label{eq:nucleus-u2}
\end{equation}
where  $A$ is a constant given by: 
\begin{equation}
A = - \frac{C_m E}{4 \pi (1 + \nu)}.  
\label{eq:nucleus-A}
\end{equation}
The elements  of the displacement vector ${\bf{u_{1}}}_{i}$ (equation \ref{eq:nucleus-u1}) are
\begin{equation}
{\bf{u_{1}}}_{i}  = 
\begin{bmatrix} 
{u_{1}}_x \\
{u_{1}}_y \\
{u_{1}}_z
\end{bmatrix}
 = \frac{A  \: (1 + \nu)}{E} \: 
\begin{bmatrix} 
\frac{\partial }{\partial x} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial }{\partial y} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial }{\partial z} {\frac{1}{{R_1}_{ij}}} 
\end{bmatrix}
 \: \Delta p_{j} \: \:dv_j \:  ,  
\label{eq:nucleus-u1-vector}
\end{equation}
where ${u_{1}}_x$, ${u_{1}}_y$ and ${u_{1}}_z$ are the $x-$, $y-$  and $z-$ components
of  ${\bf{u_{1}}}_{i} $ that gives the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus in the infinite space.

@@

We parametrized the reservoir as a grid of juxtaposed right rectangular prisms.
Each grid prism has homogeneous pressure contrasts; however, pressure variations among different prisms are allowed. 
To calculate the displacement due to the pressure change in the whole reservoir with this discretization model, we use the solution deduced for a single nucleus of strain in a homogeneous elastic semi-infinite medium (subsection \ref{solution-nucleus}) in the following way. 
First, we assume that the coordinates of the $j$th prism center are the coordinates of a nucleus of strain.
Next, the displacement field calculated at the $i$th coordinates  $(x_i$, $y_i, z_i)$  due to the pressure contrast of the $j$th prism is calculated with a integration over its volume.
Then, from equation \ref{eq:nucleus-ui_total}, the displacement field produced by the $jth$ prism can be written as 
\begin{equation}
{\bf{u}}_{i} (x_i,y_i,z_i) =  
\int_{zo_j - \Delta z/2}^{zo_j + \Delta z/2} \:\:
\int_{yo_j - \Delta y/2}^{yo_j + \Delta y/2} \:\: 
\int_{xo_j - \Delta x/2}^{xo_j + \Delta x/2} 
{\bf{u_{1}}}_{i} \: \:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j 
\: + \:
\int_{- zo_j - \Delta z/2}^{- zo_j + \Delta z/2} \:\:
\int_{yo_j - \Delta y/2}^{yo_j + \Delta y/2} \:\: 
\int_{xo_j - \Delta x/2}^{xo_j + \Delta x/2} 
{\bf{u_{2}}}_{i} \:\:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j.
\label{eq:nucleus-ui_jth_prism}
\end{equation}

In equation \ref{eq:nucleus-ui_jth_prism},  ${\bf{u_{1}}}_{i}$ is the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus in the infinite space (equations \ref{eq:nucleus-u1} and  \ref{eq:nucleus-u1-vector}),  $xo_j, yo_j$ and $zo_j$ are, respectively, the $x-, y-$ and $z-$coordinates of the $j$th prism center and $\Delta x, \: \Delta y$ and $\Delta z$ are the dimensions of the prisms along the $x-, y-$ and $z-$directions, respectively. 
Additionally, ${\bf{u_{2}}}_{i}$ is the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the effect of an image nucleus (equations \ref{eq:nucleus-u2} and  \ref{eq:nucleus-u2-vector}), where $xo_j, yo_j$ and $- zo_j$ are, respectively, the $x-, y-$ and $z-$coordinates of the $j$th image nucleus.

>>>>

\title{DisReserv (v1.0): Analytical displacement solution due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes}

@@

\Author[1]{Valeria}{C. F. Barbosa}
\Author[1]{Vanderlei}{C. Oliveira Jr.}
\Author[1,2]{Andre}{D. Arelaro}

@@

\correspondence{Valeria C. F. Barbosa (valcris@on.br)}

@@

\runningauthor{Barbosa, V. C. F., Oliveira Jr., V. C. and Arelaro, A. D.}

@@

%%\twocolumn
\begin{abstract}
We have presented analytical solutions for the displacement due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes.
These solutions are based on the similarity between the gravitational potential yielded by  a volume source under a density variation and the displacement field yielded by a volume source in a half-space under a pressure variation.
This similarity enables the use of closed expressions of the gravitational potential and its derivatives for calculating the displacement field due to a volume source under a pressure variation.
We discretized the reservoir as a grid of 3D right rectangular prisms  juxtaposed in the horizontal and vertical directions.
Each prism has homogeneous pressure; however, pressure variations among different prisms are allowed. 
This parametrization of the reservoir yields a piecewise-constant distribution of pressure  in the subsurface. 
The discrete reservoir modeling to calculate the displacement field due to this pressure variation  follows the nucleus-of-strain concept in which the center of each prism represents the coordinate of a nucleus of strain.
The displacement due to a nucleus of strain is considered the infinitesimal element of the 
displacement due to an infinitesimal reservoir.
The displacement field due to the pressure of each prism is calculated by integrating the infinitesimal element of the displacement over the volume of the prism.
Finally, the displacement due to the reservoir can be approximated by the sum of the contributions of each prism of the discretized model.
We provide python codes (DisReserv) to calculate the displacement fields due to a reservoir with arbitrary shape and distribution of pressure changes. 
The displacement field is calculated inside and outside the reservoir. 
We calculate the displacement fields in three scenarios: i) cylindrical reservoir under uniform depletion; ii) cylindrical reservoir under non uniform depletion and iii) reservoir with arbitrary geometry and under arbitrary pressure distribution.
By calculating the  stress field at the free surface, we verify that the zero stress condition is satisfied.
\end{abstract}

\copyrightstatement{}

@@

The possibility of occurrence of catastrophic events, related to subsidence due to extraction or injection of fluids in reservoir, has stimulated efforts to  develop analytical methods for modeling the displacement and stress fields due to reservoir compactation.
The physical foundation of the displacement, stress and strain fields in the subsurface due to a reduction of pressure in the reservoir comes from the theory of thermoelasticity.
Theory of thermoelasticity has been laid in the first half of the nineteenth century to describe the interaction between the thermal field and elastic bodies.
In the uncoupled thermoelasticity theory for quasi-static problems (i.e., problems with negligible inertia effects), \cite{Goodier37} employed the method of superposition using
displacement potential functions and introduced the  concept of nucleus of thermoelastic strain in an infinite space.
Specifically, Goodier’s (\citeyear{Goodier37}) method simplified the thermoelastic problem by replacing it by an isothermal elastic problem with different boundary conditions together with the solution of a Poisson’s equation \citep{Tao71}. 
\cite{Mindlin&Cheng50} and \cite{Sen51} extended the Goodier's method to a homogenous half-space.
\cite{Sharma56} deduced the displacement and stress fields in an infinite elastic plate due to a nucleus of thermoelastic strain located at a point inside it by using infinite integrals involving Bessel functions.

@@

\cite{Geertsma&Opstal73} applied the nucleus-of-strain concept in the half-space 
to calculate the spatial subsidence distribution due to the production of reservoir with an arbitrary 3D shape.
By assuming a producing reservoir embedded in an a homogeneous, isotropic, and elastic medium, and a reservoir model in which the pressure perturbations are related to the displacement field by a linear relationship, \cite{Geertsma&Opstal73} discretized the reservoir into a grid of  prisms and calculated the displacement due to the pressure change in the whole reservoir  by the superposition of the displacement due to the constant pressure change in each prism.
\cite{Tempone10} adopted the same reservoir model used in \cite{Geertsma&Opstal73} and 
extended the nucleus-of-strain concept in the half-space to consider the effects of a rigid basement. 
Similarly, \cite{Tempone10} assumed a reservoir embedded in an a homogeneous, isotropic, and elastic medium and calculated the displacement, stress and strain fields subject to uniform depletion.
The main drawbacks in \cite{Geertsma&Opstal73} and \cite{Tempone10} are the assumption of homogeneous reservoir and the solution is only valid outside the reservoir.
In these case, the displacements within the reservoir are calculated by a linear interpolation of the displacements at the upper and lower edges of the reservoir \citep{Tempone12}.

Considering an inhomogeneous poroelastic model consists of layered stratigraphy, \cite{Mehrabian&Abousleiman15} developed closed-form formulae for the  displacement and stress  fields outside and inside of the reservoir embedded within elastic strata with different mechanical properties and subjected to pore pressure disturbances due to fluid extraction or injection. 
By assuming a linear elastic semi-infinite medium, \cite{Munoz&Roehl17} developed analytical solution for the  displacement field  outside and inside of an arbitrarily-shaped reservoir under arbitrary distribution of pressure changes. 
\cite{Munoz&Roehl17} parametrized the reservoir into a grid of 3D prisms and used the nucleus-of-strain concept. 
The nucleus of strain is taking as an infinitesimal volume element  for each prism and the displacement solution due to each prism is obtained by a three-dimensional integration over the prism volume. 
The displacement field outside and inside of the reservoir is given by the summation of the displacement fields of the displacements produced by all prisms setting up the reservoir model.

The present work assumes a linear elastic semi-infinite medium and provides an analytical solution for displacement field due to an arbitrarily-shaped reservoir under arbitrary distribution of pressure changes. 
Like \cite{Munoz&Roehl17} we used the nucleus-of-strain concept and discretized the
reservoir into a grid of 3D prisms along the $x-$, $y-$ and $-z$directions.
We also consider the nucleus of strain as an infinitesimal volume element  for each prism and the displacement solution due to each prism is obtained by a three-dimensional integration over the prism volume. 
The final displacement field due to the whole reservoir is the sum of the displacements produced by the prisms. 
In contrast with \cite{Munoz&Roehl17}'s method we take advantage the similarity between 
the  equations for calculating the displacement field due to a volume source in a half-space under a pressure variation and the gravitational potential due to a volume source under a density variation.
This similarity makes possible the use of closed expressions of the gravitational potential and its derivatives produced by the 3D right rectangular prism derived by \cite{Nagyetal2000} and (\citeyear{Nagyetal2002}) and \cite{Fukushima2020}  for calculating the displacement field due to a volume source under a pressure variation.
The adopted exact analytical formulae of the gravitational field  are valid expressions either outside or inside the prisms because the implemented expressions make use of  modified arctangent function proposed by \cite{Fukushima2020}. 
We present routines written in Python language (Python 3.7.6)  to calculate the displacement fields due to a reservoir with arbitrary shape and distribution of pressure changes. 
We validate our equations by verifying that at the free surface the stress fields are null.
Tests with synthetic data validate our approach.

@@

Here, we use a Cartesian coordinate system with the $x-$axis pointing to north, the $y-$axis pointing to east and the $z-$axis pointing downward.
By considering the discrete form of equation \ref{eq:displacement-solution}, the displacement field  $ {\bf{u}}_{i} \equiv  {\bf{u}}(x_{i}, y_{i}, z_{i})$ at an arbitrary point $(x_{i}, y_{i}, z_{i})$ due to the $j$th nucleus of strain at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ will be calculated by
\begin{equation}
{\bf{u}}_{i}  = {\bf{\nabla}} \: 
{\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \: + \: {\bf{{\nabla}_{2}}} {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}, 
\label{eq:displacement_ui}
\end{equation}
In equation \ref{eq:displacement_ui}, the functions 
$\phi_{1} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}$ and  
$\phi_{2} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, -z^{\prime}_{j})}$
are, respectively, given by 
\begin{equation}
\phi_{1} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_1}_{ij} }
\label{eq:nucleus-phi1}
\end{equation}
and
\begin{equation}
\phi_{2} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_2}_{ij} }.
\label{eq:nucleus-phi2}
\end{equation}

@@

In equation \ref{eq:nucleus-phi1}, $ {R_1}_{ij}$ is the distance from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th coordinate of the nucleus of strain $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$, i.e.:
\begin{equation}
{R_1}_{ij} = {\sqrt{(x_{i}- x^{\prime}_{j})^{2} + (y_{i} - y^{\prime}_{j})^{2} + 
(z_{i} - z^{\prime}_{j})^{2}}}.
\label{eq:R1}
\end{equation}
In equation \ref{eq:nucleus-phi2}, $ {R_2}_{ij}$ is the distance from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th coordinate of the image nucleus $(x^{\prime}_{j}, y^{\prime}_{j}, - z^{\prime}_{j})$, i.e.:
\begin{equation}
{R_2}_{ij} = {\sqrt{(x_{i}- x^{\prime}_{j})^{2} + (y_{i} - y^{\prime}_{j})^{2} + 
(z_{i} + z^{\prime}_{j})^{2}}}.
\label{eq:R2}
\end{equation}
Figure \ref{fig:nucleus_strain} shows a schematic representation of the geometry
of the nucleus of strain problem in a semi-infinite medium. 
The $j$th nucleus of strain is located at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$. 
The $j$th image nucleus is located at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, - z^{\prime}_{j})$.
The distances from the $i$th coordinate point of the displacement $ (x_{i}, y_{i}, z_{i})$ to the $j$th nucleus of strain and to the $j$th image nucleus are, respectively,  
$ {R_1}_{ij}$ (equation \ref{eq:R1})  and $ {R_2}_{ij}$ (equation \ref{eq:R2}).
The free surface is a horizontal plane where the components of the stress are null.

Following the discrete form of the displacement solution (equation \ref{eq:displacement_ui}), 
the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus at the coordinates $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ can be written as:
\begin{equation}
{\bf{u}}_{i} (x_i,y_i,z_i) = {\bf{u_{1}}}_{i} (x_i,y_i,z_i) + {\bf{u_{2}}}_{i} (x_i,y_i,z_i), 
\label{eq:nucleus-ui_total}
\end{equation}
where ${\bf{u_{1}}}_{i} (x_i,y_i,z_i) \equiv {\bf{u_{1}}}_{i} $ is the gradient of the function $\phi_1$ (equation \ref{eq:nucleus-phi1}) 
\begin{equation}
{\bf{u_{1}}}_{i}  = {\bf{\nabla}} \: {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \:
= \frac{A  \: (1 + \nu)}{E} \: {\bf{\nabla}} \bigg( {\frac{1}{{R_1}_{ij}}} \bigg) 
\: \Delta p_{j} \: \:dv_j
\label{eq:nucleus-u1}
\end{equation}
and ${\bf{u_{2}}}_{i} (x_i,y_i,z_i) \equiv {\bf{u_{2}}}_{i} $ is obtained by  applying the operator ${\bf{{\nabla}_{2}}}$ (equation \ref{eq:nabla2}) to the imagem potential $\phi_2$ (equation \ref{eq:nucleus-phi2})
\begin{equation}
{\bf{u_{2}}}_{i} = {\bf{{\nabla}_{2}}} \: {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, -z^{\prime}_{j})} \:  
= \frac{A  \: (1 + \nu)}{E} \: 
\bigg[ 
(3 - 4\nu){\bf{\nabla}} \bigg(  {\frac{1}{{R_2}_{ij}}} \bigg)     
+ 2 {\bf{\nabla}} \bigg( z \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}} \bigg)  
- 4(1- \nu){\bf{\hat{z}}} {\bf{{\nabla}^{2}}}  \bigg( {\frac{z}{{R_2}_{ij}}} \bigg)
\bigg]   \: \Delta p_{j}  \:dv_j , 
\label{eq:nucleus-u2}
\end{equation}
where  $A$ is a constant given by: 
\begin{equation}
A = - \frac{C_m E}{4 \pi (1 + \nu)}.  
\label{eq:nucleus-A}
\end{equation}
The elements  of the displacement vector ${\bf{u_{1}}}_{i}$ (equation \ref{eq:nucleus-u1}) are
\begin{equation}
{\bf{u_{1}}}_{i}  = 
\begin{bmatrix} 
{u_{1}}_x \\
{u_{1}}_y \\
{u_{1}}_z
\end{bmatrix}
 = \frac{A  \: (1 + \nu)}{E} \: 
\begin{bmatrix} 
\frac{\partial }{\partial x} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial }{\partial y} {\frac{1}{{R_1}_{ij}}}  \\
\frac{\partial }{\partial z} {\frac{1}{{R_1}_{ij}}} 
\end{bmatrix}
 \: \Delta p_{j} \: \:dv_j \:  ,  
\label{eq:nucleus-u1-vector}
\end{equation}
where ${u_{1}}_x$, ${u_{1}}_y$ and ${u_{1}}_z$ are the $x-$, $y-$  and $z-$ components
of  ${\bf{u_{1}}}_{i} $ that gives the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus in the infinite space.

@@

We parameterized the reservoir as a grid of juxtaposed right rectangular prisms.
Each grid prism has homogeneous pressure contrasts; however, pressure variations among different prisms are allowed. 
To calculate the displacement due to the pressure change in the whole reservoir with this discretization model, we use the solution deduced for a single nucleus of strain in a homogeneous elastic semi-infinite medium (subsection \ref{solution-nucleus}) in the following way. 
First, we assume that the coordinates of the $j$th prism center are the coordinates of a nucleus of strain.
Next, the displacement field calculated at the $i$th coordinates  $(x_i$, $y_i, z_i)$  due to the pressure contrast of the $j$th prism is calculated with a integration over its volume.
Then, from equation \ref{eq:nucleus-ui_total}, the displacement field produced by the $jth$ prism can be written as 
\begin{equation}
{\bf{u}}_{i} (x_i,y_i,z_i) =  
\int_{z_1}^{z_2} \:\:
\int_{y_1}^{y_2} \:\: 
\int_{x_1}^{x_2} 
{\bf{u_{1}}}_{i} \: \:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j 
\: + \:
\int_{z_1}^{z_2} \:\:
\int_{y_1}^{y_2} \:\: 
\int_{x_1}^{x_2} 
{\bf{u_{2}}}_{i} \:\:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j.
\label{eq:nucleus-ui_jth_prism}
\end{equation}

In equation \ref{eq:nucleus-ui_jth_prism},  ${\bf{u_{1}}}_{i}$ is the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the $j$th single nucleus in the infinite space (equations \ref{eq:nucleus-u1} and  \ref{eq:nucleus-u1-vector}), the limits of the integrals represent the borders of the $j$th prism modeling the reservoir in the following way: $x_1$ and $x_2$ are their south and north borders; $y_1$ and  $y_2$ are their  west and east borders;  and $z_1$ and $z_2$ are their depths to the top and bottom.
Additionally, ${\bf{u_{2}}}_{i}$ is the displacement field at the coordinates $x_i$, $y_i$  and $z_i$  due to the effect of an image nucleus (equations \ref{eq:nucleus-u2} and  \ref{eq:nucleus-u2-vector}).
